Efficient evaluation of threshold queries of derived fields in a numerical simulation database

### 2014-12-29 paper.tex
Addressing reviewers' comments. Added challenges and contributions in introduction. Clarified 10^6 limitation for result sets. Added discussion of sampling approaches in related work section.

\usepackage{subfigure}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{varwidth}
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{multirow}
\usepackage{array}
%\usepackage{natbib}% bold math
\newcommand{\ddx}[2]{ \frac{ \partial #1}{ \partial {#2} } }
%\newcommand{\ddt}[1]{ \frac{\partial #1}{\partial t} }
\newcommand{\ddt}[1]{\partial_t #1}
\newcommand{\Grad}[1]{\nabla #1}
\newcommand{\Div}[1]{\Grad \cdot #1}
\newcommand{\Lap}[1]{\Delta #1}
\definecolor{light-gray}{gray}{0.95}

\lstdefinestyle{cxx}{ %
language=C++,
basicstyle=\footnotesize\ttfamily,  % the size of the fonts that are used for the code
% numbers=left,                   % where to put the line-numbers
% numberstyle=\footnotesize,      % the size of the fonts that are used for the line-numbers
% stepnumber=2,                   % the step between two line-numbers. If it's 1 each line will be numbered
% numbersep=5pt,                  % how far the line-numbers are from the code
% backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
backgroundcolor=\color{light-gray},  % choose the background color. You must add \usepackage{color}
showspaces=false,               % show spaces adding particular underscores
commentstyle=\color{cyan},
stringstyle=\ttfamily\color{red},
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
frame=single,                 % adds a frame around the code
tabsize=2,                      % sets default tabsize to 2 spaces
captionpos=b,                   % sets the caption-position to bottom
breaklines=true,                % sets automatic line breaking
breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
title=\lstname,                 % show the filename of files included with \lstinputlisting; also try caption instead of title
% morekeywords={char,double},
escapeinside={\%*}{*)},          % if you want to add a comment within your code
% morekeywords={*,...}          % if you want to add more keywords to the set
% morekeywords={char,double}    % if you want to add more keywords to the set
frameround=fttt
}

\lstdefinestyle{fortran}{ %
language=Fortran,
basicstyle=\footnotesize\ttfamily,  % the size of the fonts that are used for the code
% numbers=left,                   % where to put the line-numbers
% numberstyle=\footnotesize,      % the size of the fonts that are used for the line-numbers
% stepnumber=2,                   % the step between two line-numbers. If it's 1 each line will be numbered
% numbersep=5pt,                  % how far the line-numbers are from the code
% backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
backgroundcolor=\color{light-gray},  % choose the background color. You must add \usepackage{color}
showspaces=false,               % show spaces adding particular underscores
commentstyle=\color{cyan},
stringstyle=\ttfamily\color{red},
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
frame=single,                 % adds a frame around the code
tabsize=2,                      % sets default tabsize to 2 spaces
captionpos=b,                   % sets the caption-position to bottom
breaklines=true,                % sets automatic line breaking
breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
title=\lstname,                 % show the filename of files included with \lstinputlisting; also try caption instead of title
% morekeywords={char,double},
escapeinside={\%*}{*)},          % if you want to add a comment within your code
% morekeywords={*,...}          % if you want to add more keywords to the set
% morekeywords={char,double}    % if you want to add more keywords to the set
frameround=fttt
}

@@

%\conferenceinfo{EuroMPI}{2013 Madrid, Spain}

\title{
Efficient evaluation of threshold queries of derived fields in a numerical simulation database
}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
\alignauthor Kalin Kanov \\
     \affaddr{Department of Computer Science} \\
     \affaddr{IDIES} \\
     \affaddr{Johns Hopkins University} \\
     \affaddr{Baltimore, MD 21218} \\
     \email{kalin@cs.jhu.edu}
\alignauthor Randal Burns \\
     \affaddr{Department of Computer Science} \\
     \affaddr{IDIES} \\
     \affaddr{Johns Hopkins University}\\
     \affaddr{Baltimore, MD 21218} \\
     \email{randal@cs.jhu.edu} 
\alignauthor Cristian C. Lalescu \\
     \affaddr{Department of Applied Mathematics and Statistics} \\
     \affaddr{IDIES} \\
     \affaddr{Johns Hopkins University} \\
     \affaddr{Baltimore, MD 21218} \\
     \email{clalesc1@jhu.edu}
}
%% \and % use '\and' if you need 'another row' of author names
%% % There's nothing stopping you putting the seventh, eighth, etc.
%% % author on the opening page (as the 'third row') but we ask,
%% % for aesthetic reasons that you place these 'additional authors'
%% % in the \additional authors block, viz.
%% \additionalauthors{Additional authors: John Smith (The Th{\o}rv{\"a}ld Group,
%% email: {\texttt{jsmith@affiliation.org}}) and Julius P.~Kumquat
%% (The Kumquat Consortium, email: {\texttt{jpkumquat@consortium.net}}).}
%% \date{30 July 1999}
%% % Just remember to make sure that the TOTAL number of authors
%% % is the number that will appear on the first page PLUS the
%% % number that will appear in the \additionalauthors section.

@@

\begin{abstract}
%Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
%The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
%interested in out of the box. Moreover, d
In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
simulation datasets stored in a cluster of relational databases. 
Such simulations are commonplace in the study of many different physical phenomena. 
The datasets produced by these simulations are in the TB and even PB ranges. 
Data-intensive computations that examine entire time-steps of the simulation data are impractical to perform
locally by the user taking days or months to iterate over the entire dataset. 
The integrated method for the evaluation of threshold queries that we have developed achieves scalability through data-parallel execution
of the computations on the nodes of the database cluster. 
We extend the scientific database cluster with the introduction of an application-aware cache for query results, building on the concept of semantic caching.
The cache has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
In the case of computational turbulence this allows scientists to quickly focus on the most intense regions in any time-step or the dataset as a whole,
which greatly speeds up the rate of scientific exploration and discovery. 
Furthermore, caching the results of threshold queries preserves both the I/O and computation effort used to obtain them, which allows subsequent
queries for some of the most interesting regions and features of the dataset to run orders of magnitude faster.
\end{abstract}

%% % A category with the (minimum) three required fields
%\category{}{Extending Data Base Technology}
%% %A category including the fourth, optional field follows...
%% \category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

@@

\section{Introduction}
%What is the problem?
Better instruments, faster and bigger supercomputers and easier collaboration and sharing of data in the sciences have introduced the need
to manage increasingly large data sets. Data-intensive systems and architectures have been developed with the goal of storing and providing
fast access to such data sets. Examples include the GrayWulf and Data-Scope clusters \cite{Szalay, DataScope} at Johns Hopkins, which have capacity 
of 1.1PB and 11PB
respectively. One of their missions is to provide persistent storage and public access to world-class numerical simulation data. 
These systems differ from the traditional HPC systems in that they aim to achieve high aggregate throughput by balancing computation
capabilities with I/O and network bandwidth. The computing systems and services developed on top of these platforms are more than pure
storage engines and usually have complex analysis routines built-in, which has largely been driven by the ``move the computation to the data"
paradigm \cite{fourthparadigm}. These built-in analysis routines are most often not novel themselves. They implement core scientific functionality for the
study of the particular scientific phenomena, which was observed or simulated in the first place. The analysis routines however require novel evaluation
strategies and methods for their execution. They have to operate on large array datasets distributed across multiple nodes of a cluster of relational
databases and in order to reduce their running times they have to make efficient use of the cluster resources.

Finding the locations or regions of highest vorticity or those with the largest norms of the velocity or other fields of interest is
functionality that enables new insights in the study of fluid dynamics. Analysis of this kind coupled with the ability to analyze time-series datasets both 
forward and backward in time has transformed our understanding of turbulence \cite{Eyink}. Furthermore, threshold, top-$k$ queries and similarity search
in general are important in many different disciplines. We introduce an efficient evaluation strategy for threshold
queries over time-series datasets stored in a cluster of relational databases. Our method evaluates not only threshold queries of the vector or 
scalar field data stored in the database, but also performs thresholding of \emph{derived} fields.

%Why is it hard? (E.g., why do naive approaches fail?)
Evaluating threshold queries within the database cluster allows scientists with modest computational and network capability to narrow down on and
examine some of the most interesting regions and features in the dataset and focus on the subsequent analysis needed to understand these events.
The evaluation of threshold queries of derived fields requires a kernel computation at each point on the grid in order to compute the field to be thresholded.
The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
In order to evaluate such a query for a given time-step, all of the data for that time-step must be read and each data point 
will appear in $N$ kernels of computation, in which $N$ is the size of the kernel. The evaluation is both data and computation intensive.
Obtaining the derived field and thresholding locally by the user requires not only the computation of the derived field over an entire time-step server-side 
but also the transfer of a large amount of data over the network, most of which are subsequently discarded. 
One of our collaborators reported that such a local evaluation of a threshold
query over an entire time-step took over 20 hours. Even when submitting several requests in parallel for the derived field that he was interested in it
would take months to iterate over the entire dataset. This highlighted the need for providing the capability through an integrated approach,
which performs the evaluation server-side.

@@

%Why is it interesting and important?
Thresholding allows scientists to obtain and examine the most intense regions and features in the dataset in the case of turbulence. 
These are often the locations that have the largest vorticity norms and have intense vortices or reconnection events. 
In magnetohydrodynamics, the locations of largest electric current are of great interest for similar reasons.
It is important that threshold queries are evaluated in an efficient manner, because often further subsequent examination
and analysis is required to understand the physics that drive these intense events.

%Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
Database systems support rollup queries, including top-$k$ queries, but in most cases these
queries apply only simple linear score functions on the attribute values of individual records.
Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning (see \cite{Ilyas} for a
survey of top-$k$ evaluation strategies).
This is an assumption that we cannot make for the functions used to compute all the different possible derived fields of interest in fluid mechanics.
Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
In contrast our approach computes the value of a \emph{derived} field at each location in the array by examining the data at all neighboring locations, which
are within the kernel of the computation. The functions used to derive the field may even be non-linear.
Finally, a top-$k$ approach may not be suitable in the cases where scientists are interested in performing threshold queries at different time-steps as the
same threshold level will produce different number of points in the result set for different time-steps.

@@

%What are the key components of my approach and results? Also include any specific limitations.
We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
stored in the database. The derived fields, which our method computes and operates on are derived by performing kernel computations over the 
stored array data.
Our method makes effective use of the cluster resources and achieves high throughput and scalability. 
We exploit the parallelism 
available in the cluster by means of data-parallel execution of the computations. Each computation is distributed across the nodes of the cluster and
is executed in parallel by several processes within a database node by creating multiple connections to each node. We extend the database management
system with an
application-aware cache for query results. We build on the idea of an application-aware cache introduced by Lopez et al. \cite{Lopez} and more broadly 
on the concept of semantic caching \cite{Dar}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.,
we cache the actual results of the computation, which leads to query performance improvement of over an order of magnitude.
The cache that we describe is therefore a limited version of a semantic cache.

@@

The method that we have developed for the evaluation of threshold queries of derived fields was deployed and integrated into the 
Johns Hopkins Turbulence Databases (JHTDB) \cite{Yi, Perlman}. The JHTDB, built on top of the GrayWulf and Data-Scope clusters, serves as a public
virtual laboratory for
the study of turbulent phenomena. The JHTDB stores several datasets, which are the output of high-resolution numerical simulations of turbulence.
The 3d time-series data are partitioned into small sub-cubes and stored in relational databases distributed across the nodes of the cluster.
Access to the data is provided by means of Web-services and a variety of analysis functions have been
implemented and can be executed through Web-service calls (Fig. \ref{fig:jhtdb_diagram}).
At present the service hosts four datasets, which are available publicly. 
The data are the output of numerical simulations of forced isotropic turbulence, magnetohydrodynamics (MHD), channel 
flow turbulence and homogenous buoyancy driven turbulence.
The total amount of space occupied by the datasets is over 230 TB.

@@

The JHTDB provide a variety of data-intensive analysis routines that are executed on the database nodes. These include interpolation, 
differentiation, particle tracking and spatial filtering. These tasks are often data-intensive and in order to leverage the capabilities of the cluster we
have developed data-driven batch processing techniques for their evaluation \cite{KanovSC11, KanovSC12}. Most of these tasks usually operate on subsets
of the space or a collection of individual target locations within a time-step. In contrast, threshold and top-$k$ queries usually have to examine the entire 
data volume of a time-step or a signification portion of it. Furthermore, the data product of threshold queries is much smaller in relation to the amount of 
data that need to be examined. This fact combined with the fact that subsequent queries can reuse 
previously computed results makes the query results suitable to caching.

@@

{\bf Distributed data-parallel execution:} In most cases threshold queries operate over an entire time-step. Each such query is subdivided by 
the mediator into queries submitted to
each of the database nodes. Each node evaluates the query over the data that it has stored locally. Only a small amount of data along the boundary
need to be requested from adjacent nodes. The size of the band of data that may not be available locally is equal to a kernel half-width. 
Such a band is needed on each of the sides of the box forming the domain of the computation.  
The data are read into memory and the particular field requested is computed at each of the locations on
the grid. The same strategy applies when utilizing multiple processes per node. The norm or absolute value of the derived field at each location is compared 
against the specified threshold and if the value is higher it is maintained along with
the spatial coordinates of the location in a list. We limit the maximum size of this list to around 1,000,000 locations in order to prevent having to return
the entire time-step or a significant fraction of it for queries with thresholds that are set too low.

@@

@@

%\kk{Cite results for top-$k$ queries in databases and distributed systems. Remark that non of these strategies operate on large array datasets 
%and they do not operate on derived fields, where each value of the derived field depends on multiple values from the original field and requires a kernel
%computation accessing multiple data points in a kernel around the target location.}

@@

\begin{algorithm}
\caption{Get points above threshold using cache}\label{alg:cache_lookup}
\begin{algorithmic}[1]
\Require Dataset $d$, Field $f$, Timestep $t$, Threshold $k$, Query box $q = [x_l,y_l,z_l,x_u,y_u,z_u]$
\Procedure{GetThreshold}{}
\State $points \gets List()$
\State $updateCache \gets false$
\State $query \gets \text{SELECT * FROM cachedb..cacheInfo}$
\Statex[4] $\text{WHERE dataset = {\it d} AND field = {\it f}}$
\Statex[4] $\text{AND timestep = {\it t}}$
\State $command \gets SqlCommand(query)$
\State $reader \gets command.ExecuteReader()$
\If {$reader.HasRows()$}
\State $k_s \gets reader[``threshold"]$ \Comment{Stored threshold}
\State $start \gets reader[``startIndex"]$
\State $end \gets reader[``endIndex"]$
\State $ordinal \gets reader[``ordinal"]$
\If {$k \le k_s \And q \in \text{[start, end]}$}
\State $\textit{query} \gets \text{SELECT * FROM cachedb..cacheData}$
\Statex[4] $\text{WHERE cacheInfoOrdinal = {\it ordinal}}$
\State $command \gets SqlCommand(query)$
\State $reader \gets command.ExecuteReader()$
\While{reader.Read()}
\State $location \gets reader[``zindex"]$
\State $norm \gets reader[``dataValue"]$
\If {$norm \ge k \And location \in q$}
\State $\textit{points.Add(new Point(location, norm))}$
\EndIf
\EndWhile
\Else
\State $updateCache \gets true$
\EndIf
\Else
\State $updateCache \gets true$
\EndIf
\If {$updateCache$}
\State $\text{Retrieve data covering {\it q} from DB.}$
\ForAll{$p \in q$}
\State{Compute $f$ at $p$.}
\If{$\| f(p) \| \ge k$}
\State $\textit{points.Add(new Point(p, $\| f(p) \|$))}$
\EndIf
\EndFor
\State{Update cacheInfo and cacheData tables.}
\EndIf
\State{\Return $points$}
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields that
have to be computed from the stored data. It is defined as the curl of the velocity field.
As described in section \ref{science_use_cases}
thresholding the vorticity field is important in the study of fluid dynamics and obtaining the locations of maximum vorticity can lead to new insights into
the development of the most intense vortices that are observed. 
%\kk{Consider showing other fields as well; possibly thresholding the velocity by itself}

@@

The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Figure
\ref{fig:cache_overhead} compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate
the cache first (blue and red bars in the figure). The execution times are also shown in Table \ref{tab:cache_effectiveness}. 
For these experiments we requested
the locations with norms of the vorticity above thresholds at different levels. We refer the reader to Figure \ref{fig:pdf}, which shows the distribution of values of
the norm of the vorticity field in the MHD dataset to get an appreciation of the different threshold values used in the experiments. 
For the first set the threshold was set high (80.0) and only $\sim$4,300 points 
(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold (60.0) was chosen and $\sim$87,000 points 
(or 0.0081\% of all points) were above the threshold. Finally, a low threshold (44.0) was chosen for the last set with $\sim$900,000 points 
(or 0.0847\% of all points) above the threshold. For each set a random
time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user.

>>>>

\usepackage{subfigure}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{varwidth}
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{multirow}
\usepackage{array}
\newcommand{\ddx}[2]{ \frac{ \partial #1}{ \partial {#2} } }
\newcommand{\ddt}[1]{\partial_t #1}
\newcommand{\Grad}[1]{\nabla #1}
\newcommand{\Div}[1]{\Grad \cdot #1}
\newcommand{\Lap}[1]{\Delta #1}
\definecolor{light-gray}{gray}{0.95}

@@

\title{
Efficient evaluation of threshold queries of derived fields in a numerical simulation database
}

\numberofauthors{3} 
\author{
\alignauthor Kalin Kanov \\
     \affaddr{Department of Computer Science} \\
     \affaddr{IDIES} \\
     \affaddr{Johns Hopkins University} \\
     \affaddr{Baltimore, MD 21218} \\
     \email{kalin@cs.jhu.edu}
\alignauthor Randal Burns \\
     \affaddr{Department of Computer Science} \\
     \affaddr{IDIES} \\
     \affaddr{Johns Hopkins University}\\
     \affaddr{Baltimore, MD 21218} \\
     \email{randal@cs.jhu.edu} 
\alignauthor Cristian C. Lalescu \\
     \affaddr{Department of Applied Mathematics and Statistics} \\
     \affaddr{IDIES} \\
     \affaddr{Johns Hopkins University} \\
     \affaddr{Baltimore, MD 21218} \\
     \email{clalesc1@jhu.edu}
}

@@

\begin{abstract}
%Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
%The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
%interested in out of the box. Moreover, d
In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
simulation datasets stored in a cluster of relational databases. 
Such simulations are commonplace in the study of many different physical phenomena. 
The datasets produced by these simulations are in the TB and even PB ranges. 
Data-intensive computations that examine entire time-steps of the simulation data are impractical to perform
locally by the user taking days or months to iterate over the entire dataset. 
The integrated method for the evaluation of threshold queries that we have developed achieves scalability through data-parallel execution
of the computations on the nodes of an analysis database cluster. 
We extend the scientific analysis environment with the introduction of an application-aware cache for query results, building on the concept of semantic caching.
The cache has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
In the case of computational turbulence this allows scientists to quickly focus on the most intense events and interesting regions in any time-step or 
the dataset as a whole, which greatly speeds up the rate of scientific exploration and discovery. 
Furthermore, caching the results of threshold queries preserves both the I/O and computation effort used to obtain them, which allows subsequent
queries for some of the most interesting regions and features of the dataset to run orders of magnitude faster.
\end{abstract}

%% % A category with the (minimum) three required fields
\category{H.2.8}{Database Management}{Database Applications -- scientific, spatial databases and GIS}
%% %A category including the fourth, optional field follows...
%% \category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

@@

\section{Introduction}
%What is the problem?
Better instruments, faster and bigger supercomputers and easier collaboration and sharing of data in the sciences have introduced the need
to manage increasingly large data sets. Data-intensive systems and architectures have been developed with the goal of storing and providing
fast access to such data sets. Examples of such analysis environments include the GrayWulf and Data-Scope clusters \cite{Szalay, DataScope} at 
Johns Hopkins, which have capacity of 1.1PB and 11PB
respectively. One of their missions is to provide persistent storage and public access to world-class numerical simulation data. 
These systems differ from the traditional HPC environments in that they aim to achieve high aggregate throughput by balancing computation
capabilities with I/O and network bandwidth. The computing systems and services developed on top of these platforms are more than pure
storage engines and usually have complex analysis routines built-in, which has largely been driven by the ``move the computation to the data"
paradigm \cite{fourthparadigm}. These built-in analysis routines are most often not novel themselves. They implement core scientific functionality for the
study of the particular scientific phenomena, which was observed or simulated in the first place. The analysis routines however require novel evaluation
strategies and methods for their execution. They have to operate on large array datasets distributed across multiple nodes of a cluster of relational
databases and in order to reduce their running times they have to make efficient use of the cluster resources.

Finding the locations or regions of highest vorticity or those with the largest norms of the velocity or other fields of interest is
functionality that enables new insights in the study of fluid dynamics. Analysis of this kind coupled with the ability to analyze time-series datasets both 
forward and backward in time has transformed our understanding of turbulence \cite{Eyink}. Furthermore, threshold, top-$k$ queries and similarity search
in general are important in many different disciplines. We introduce an efficient evaluation strategy for threshold
queries over time-series datasets stored in a cluster of relational databases. Our method evaluates not only threshold queries of the vector or 
scalar field data stored in the database, but also performs thresholding of \emph{derived} fields. The main challenge that our approach tackles is that the
data-intensive computation of \emph{derived} fields has to be carried out on-demand for extremely large array datasets stored in an analysis cluster environment 
comprised of multiple database nodes. Here we focus on the evaluation of threshold queries over the fields derived from the data stored in the cluster as
these queries are the most interesting scientifically. However, our approach applies to the evaluation of top-$k$ queries, rollup queries and data-reducing 
queries in general.

%Why is it hard? (E.g., why do naive approaches fail?)
Evaluating threshold queries within the database cluster allows scientists with modest computational and network capability to narrow down on and
examine some of the most interesting regions and features in the dataset and focus on the subsequent analysis needed to understand these events.
The evaluation of threshold queries of derived fields requires a stencil kernel computation at each point on the grid in order to compute the field to be thresholded.
The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
In order to evaluate such a query for a given time-step, all of the data for that time-step must be read and each data point 
will appear in $N$ kernels of computation, in which $N$ is the size of the kernel. The evaluation is both data and computation intensive.
Obtaining the derived field and thresholding locally by the user requires not only the computation of the derived field over an entire time-step server-side 
but also the transfer of a large amount of data over the network, most of which are subsequently discarded. 
One of our collaborators reported that such a local evaluation of a threshold
query over an entire time-step took over 20 hours. Even when submitting several requests in parallel for the derived field that he was interested in it
would take months to iterate over the entire dataset. This highlighted the need for providing the capability through an integrated approach,
which performs the evaluation server-side.

@@

%Why is it interesting and important?
Thresholding allows scientists to obtain and examine the regions containing the most intense events and features in the dataset in the case of turbulence. 
These are often the locations that have the largest vorticity norms and have intense vortices or reconnection events. 
In magnetohydrodynamics, the locations of largest electric current are of great interest for similar reasons.
It is important that threshold queries are evaluated in an efficient manner, because often further subsequent examination
and analysis is required to understand the physics that drive these intense events.

%Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
There are several challenges that arise during the evaluation of threshold queries of derived fields in an analysis database cluster. The field variables have to be
evaluated \emph{on-demand} from the array data stored in the database cluster. The computations are data-intensive as they apply stencil kernels on extremely
large multidimensional array datasets. Such a stencil kernel computation has to be performed at each location on the grid as opposed to at a particular number 
of target locations. Finally, the evaluation needs to be distributed across the nodes of the database cluster to avoid the unnecessary movement of data across 
the network and to achieve scalability.

@@

%What are the key components of my approach and results? Also include any specific limitations.
We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
stored in the database. The derived fields, which our method computes and operates on are derived by performing stencil kernel computations over the 
stored array data.
Our method makes effective use of the cluster resources and achieves high throughput and scalability. 
We exploit the parallelism 
available in the cluster by means of data-parallel execution of the computations. Each computation is distributed across the nodes of the cluster and
is executed in parallel by several processes within a database node by creating multiple connections to each node. We extend the database management
system with an
application-aware cache for query results. We build on the idea of an application-aware cache introduced by Lopez et al. \cite{Lopez} and more broadly 
on the concept of semantic caching \cite{Dar}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.,
we cache the actual results of the computation, which leads to query performance improvement of over an order of magnitude.
The cache that we describe is therefore a limited version of a semantic cache.

The contributions of this paper are the following:
\begin{itemize}
\item Computing derived fields of large simulation data on-demand and evaluating threshold queries on them at extreme scale. For the first time providing large
data analytics capabilities that examine entire time-steps of the simulation transparently to the user in a production analysis environment.
\item Achieving this through the combination of existing data management techniques such as data parallelism and semantic caching and taking advantage of heterogeneous architectures (sharded database cluster with several SSDs per node).
\item Evaluating the proposed method on data-intensive workloads in a live production environment and showing scalability results on datasets hundreds of terabytes in size.
\end{itemize}

@@

The method that we have developed for the evaluation of threshold queries of derived fields was deployed and integrated into the 
Johns Hopkins Turbulence Databases (JHTDB) \cite{Yi, Perlman}. It solves a pressing problem in a production scientific analysis environment, which differs 
from the traditional supercomputing environments and provides large data analytics capabilities transparently to the public. 
The JHTDB, built on top of the GrayWulf and Data-Scope clusters, serves as a public virtual laboratory for
the study of turbulent phenomena. The JHTDB stores several datasets, which are the output of high-resolution numerical simulations of turbulence.
The 3d time-series data are partitioned into small sub-cubes and stored in relational databases distributed across the nodes of the cluster.
Access to the data is provided by means of Web-services and a variety of analysis functions have been
implemented and can be executed through Web-service calls (Fig. \ref{fig:jhtdb_diagram}).
At present the service hosts four datasets, which are available publicly. 
The data are the output of numerical simulations of forced isotropic turbulence, magnetohydrodynamics (MHD), channel 
flow turbulence and homogenous buoyancy driven turbulence.
The total amount of space occupied by the datasets is over 230 TB.

@@

The JHTDB provides a variety of data-intensive analysis routines that are executed on the database nodes. These include interpolation, 
differentiation, particle tracking and spatial filtering. These tasks are often data-intensive and in order to leverage the capabilities of the cluster we
have developed data-driven batch processing techniques for their evaluation \cite{KanovSC11, KanovSC12}. Most of these tasks usually operate on subsets
of the space or a collection of individual target locations within a time-step. In contrast, threshold and top-$k$ queries usually have to examine the entire 
data volume of a time-step or a signification portion of it. Furthermore, the data product of threshold queries is much smaller in relation to the amount of 
data that need to be examined. This fact combined with the fact that subsequent queries can reuse 
previously computed results makes the query results suitable to caching.

@@

{\bf Distributed data-parallel execution:} In most cases threshold queries operate over an entire time-step. Each such query is subdivided by 
the mediator into queries submitted to
each of the database nodes. Each node evaluates the query over the data that it has stored locally. Only a small amount of data along the boundary
need to be requested from adjacent nodes. The size of the band of data that may not be available locally is equal to a kernel half-width. 
Such a band is needed on each of the sides of the box forming the domain of the computation.  
The data are read into memory and the particular field requested is computed at each of the locations on
the grid. The same strategy applies when utilizing multiple processes per node. The norm or absolute value of the derived field at each location is compared 
against the specified threshold and if the value is higher it is maintained along with
the spatial coordinates of the location in a list.

@@

\begin{algorithm}
\caption{Get points above threshold using cache}\label{alg:cache_lookup}
\begin{algorithmic}[1]
\Require Dataset $d$, Field $f$, Timestep $t$, Threshold $k$, Query box $q = [x_l,y_l,z_l,x_u,y_u,z_u]$
\Procedure{GetThreshold}{}
\State $points \gets List()$
\State $updateCache \gets false$
\State $query \gets \text{SELECT * FROM cachedb..cacheInfo}$
\Statex[4] $\text{WHERE dataset = {\it d} AND field = {\it f}}$
\Statex[4] $\text{AND timestep = {\it t}}$
\State $command \gets SqlCommand(query)$
\State $reader \gets command.ExecuteReader()$
\If {$reader.HasRows()$}
\State $k_s \gets reader[``threshold"]$ \Comment{Stored threshold}
\State $start \gets reader[``startIndex"]$
\State $end \gets reader[``endIndex"]$
\State $ordinal \gets reader[``ordinal"]$
\If {$k \ge k_s \And q \in \text{[start, end]}$}
\State $\textit{query} \gets \text{SELECT * FROM cachedb..cacheData}$
\Statex[4] $\text{WHERE cacheInfoOrdinal = {\it ordinal}}$
\State $command \gets SqlCommand(query)$
\State $reader \gets command.ExecuteReader()$
\While{reader.Read()}
\State $location \gets reader[``zindex"]$
\State $norm \gets reader[``dataValue"]$
\If {$norm \ge k \And location \in q$}
\State $\textit{points.Add(new Point(location, norm))}$
\EndIf
\EndWhile
\Else
\State $updateCache \gets true$
\EndIf
\Else
\State $updateCache \gets true$
\EndIf
\If {$updateCache$}
\State $\text{Retrieve data covering {\it q} from DB.}$
\ForAll{$p \in q$}
\State{Compute $f$ at $p$.}
\If{$\| f(p) \| \ge k$}
\State $\textit{points.Add(new Point(p, $\| f(p) \|$))}$
\EndIf
\EndFor
\State{Update cacheInfo and cacheData tables.}
\EndIf
\State{\Return $points$}
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

We impose a limit on the maximum number of locations that can be returned as a result of a threshold query in order to prevent having to return
the entire time-step or a significant fraction of it for queries with thresholds that are set too low. Currently this limit is around 1,000,000 locations, which is
sufficient to examine a time-step in detail while limiting the amount of data that have to be returned to the user over the network as well as the amount of
data that have to be cached. Users receive an error message notifying them if their request has a threshold that is set too low. If a user
is interested in obtaining more data he or she can request the values of the derived field directly. Alternatively, if they are interested in the density distribution
of values they can examine the probability density function (e.g. Fig. \ref{fig:pdf}), which is computed using a similar strategy to threshold queries.

@@

@@

For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields that
have to be computed from the stored data. It is defined as the curl of the velocity field.
As described in section \ref{science_use_cases}
thresholding the vorticity field is important in the study of fluid dynamics and obtaining the locations of maximum vorticity can lead to new insights into
the development of the most intense vortices observed in the dataset.

@@

The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Figure
\ref{fig:cache_overhead} compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate
the cache first (blue and red bars in the figure). The execution times are also shown in Table \ref{tab:cache_effectiveness}. 
For these experiments we requested
the locations with norms of the vorticity above thresholds at different levels. We refer the reader to Figure \ref{fig:pdf}, which shows the distribution of values of
the norm of the vorticity field in the MHD dataset to get an appreciation of the different threshold values used in the experiments. 
For the first set the threshold was set high (80.0) and only $\sim$4,300 points 
(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold (60.0) was chosen and $\sim$87,000 points 
(or 0.0081\% of all points) were above the threshold. Finally, a low threshold (44.0) was chosen for the last set and there were $\sim$900,000 points 
(or 0.0847\% of all points) above the threshold. For each set a random
time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user.
